Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning

### 2022-08-14 manuscript/manuscript.Rmd

Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
This paper is not intended to give a complete answer to whether deep learning is the best suited method for tipping point forecasting problems as this will take numerous studies to resolve; instead, this paper aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems.
Given the difficulty of forecasting never-before-observed behavior, as illustrated by the Hopf and saddle-node bifurcation scenarios, there is good reason for research to focus more on the kind of qualitative predictions long emphasized in the literature on early warning signals and resilience [@scheffer2015].
Recently, ML techniques developed for classification rather than the ML methods used in regression and forecasting models considered here have demonstrated a more nuanced ability to reliably detect different classes of critical transitions in time-series data [@bury2021; @lapeyrolerie2021].
Rather than seeking to provide managers with quantitative, probabilistic forecasts reflecting a broad uncertainty in possible outcomes, this literature has sought to emphasize only a more qualitative form of prediction, such as establishing whether a system is either "resilient" or "approaching a critical transition."
Decision sciences have long emphasized the importance of reconciling the qualitative predictions of resilience thinking with quantitative forecasts of future states [@folke2009; @polasky2011].
Such approaches could be valuable in concert with probabilistic forecasts considered here, providing a possible mechanism to identify when the probabilistic forecast is least reliable.

>>>>

Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
This paper is not intended to give a complete answer to whether deep learning is the best suited method for tipping point forecasting problems as this will take numerous studies to resolve; instead, this paper aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems.
Given the difficulty of forecasting never-before-observed behavior, as illustrated by the Hopf and saddle-node bifurcation scenarios, there is good reason for research to focus more on the kind of qualitative predictions long emphasized in the literature on early warning signals and resilience [@scheffer2015].
Recently, ML techniques developed for classification rather than the ML methods used in regression and forecasting models considered here have demonstrated a more nuanced ability to reliably detect different classes of critical transitions in time-series data [@bury2021; @lapeyrolerie2021].
Rather than seeking to provide managers with quantitative, probabilistic forecasts reflecting a broad uncertainty in possible outcomes, this literature has sought to emphasize only a more qualitative form of prediction, such as establishing whether a system is either "resilient" or "approaching a critical transition."
Decision sciences have long emphasized the importance of reconciling the qualitative predictions of resilience thinking with quantitative forecasts of future states [@fischer2009; @polasky2011].
Such approaches could be valuable in concert with probabilistic forecasts considered here, providing a possible mechanism to identify when the probabilistic forecast is least reliable.
